%% LaTeX packages such as hyperref
\usepackage{hyperref}

%% science LaTeX definitions
\newcommand\hGpc{\mbox{$h^{-1}$\,Gpc}}

%% journal abbreviations that might be needed
\providecommand\aap{Astronomy \& Astrophysics}
\providecommand\cqg{Classical and Quantum Gravity}
\providecommand\BASI{Bulletin of the Astronomical Society of India}
\providecommand\AmJPhys{American Journal of Physics}
\providecommand\FoundPhys{Foundations of Physics}
\providecommand\jrasc{JRASC}
\providecommand\jrasc{J. Roy. Ast. Soc. Canada}
%
\providecommand\pasj{PASJ}
\providecommand\pasa{PASA}
\providecommand\pasa{Proc.~Astr.~Soc.~Austr.}
\providecommand\cqg{CQG}
\providecommand\fcp{Fund.~Cosmic~Phys.}
\providecommand\jcap{JCAP}
\providecommand\BASI{Bull. Astr. Soc. India}
\providecommand\AmJPhys{Am. J. Phys.}
\providecommand\FoundPhys{Found. Phys.}
\providecommand\jams{J.~Amer.~Math.~Soc.}
\providecommand\jetpL{Journal of Experimental and Theoretical Physics Letters}
\providecommand\PRL{Physical Review Letters}
\providecommand\prd{Physical Review D}
\providecommand\PRD{\prd}
\providecommand\APPB{Acta Physica Polonica B}
\providecommand\APPBPS{Acta Physica Polonica B Proceedings Supplement}
\providecommand\ijmpd{International Journal of Modern Physics D}
%
\providecommand\advspaceres{Advances in Space Research}
\providecommand\physrep{Phys.Rep.}
\providecommand\grg{General Relativity and Gravitation}
\providecommand\na{New Astronomy}
\providecommand\ISBN{ISBN:~}
\providecommand\jsiam{Journal of the Society for Industrial and Applied Mathematics}
\providecommand\jsiam{J.~Soc.~Indust.~App.~Math.}
%
\providecommand\apj{Astrophys.J.}
\providecommand\apjl{Astrophys.J.Lett.}
\providecommand\apjs{Astrophys.J.Supp.}
\providecommand\ssr{Space~Sci.~Rev.}
\providecommand\annalesBruxelles{Annales de la Soci\'et\'e Scientifique de Bruxelles}
\providecommand\annalesBruxelles{Ann.~Soc.~Sci~Brux.}
\providecommand\nat{Nature}
\providecommand\pnas{Proceedings of the National Academy of Sciences}
\providecommand\pnas{Proc.~Nat.~Acad.~Sci.}
\providecommand\commmathphys{Communications in Mathematical Physics}
\providecommand\commmathphys{Commun.~Math.~Physics}
\providecommand\annrevnucpartphys{Annual Review of Nuclear and Particle Science}
\providecommand\annrevnucpartphys{Ann.~Rev.~Nucl.~Part.~Sci.}
\providecommand\nucphysbprocsupp{Nuclear Physics B Proceedings Supplements}
\providecommand\nucphysbprocsupp{Nucl.~Phys.~B~Proc.~Supp.}
\providecommand\SPIEConfSeries{Society of Photo-Optical Instrumentation Engineers (SPIE) Conference Series}
\providecommand\SPIEConfSeries{SPIE Conf. Ser. }
\providecommand\scandjstat{Scand.~J.~Stat.}
\providecommand\annmathstat{Ann.~Math.~Stat.}
\providecommand\newjphys{New Journal of Physics}
